Phylogenetics and the human microbiome.

### 2014-06-24 sbreview.tex
Sorting references.

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that a collection of sequences need not be representative of the communities from which they were derived \citep{morgan2010metagenomic}.
Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
This can distort inferences concerning actual microbe abundances based on read copy number.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts method \citep{felsenstein1985phylogenies} to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{case2007rpob,mcnabb2004hsp65}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except in the inference of operational taxonomic units as described below.

@@

Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{greenblum2012metagenomic,abubucker2012metabolic}.
Discussing these methods is beyond the scope of this paper, as is the sequencing of mRNA in bulk which is called ``metatranscriptomics".

\subsection{Whole genomes}
In addition to conventional genome sequencing for microbial genomes, whole-genome sequencing from culture is currently being used for microbial outbreak tracking \citep{koser2012rapid,snitkin2012tracking}.
The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genomes sampled from the environment and grown in culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
This data may become more common for unculturable organisms as single-cell sequencing methods improve \citep[reviewed in][]{kalisky2011single}.
The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{howe2012assembling,pell2012scaling,iverson2012untangling,emerson2012metagenomic,podell2013assembly}.

@@

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on counts of DNA substrings of a given length, which are often called $k$-mer classifiers \citep{wang2007naive,rosen2008metagenome}.
For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the taxonomic tree.
\citet{munch2008statistical,munch2008fast} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \citet{Koslicki2013-mc}, and \citet{Droge2014-lj}.
\citet{treangen2013metamos} report shorter run times and much higher accuracy (again, for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.

@@

There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a so-called fixed reference tree, sometimes with uncertainty estimates.
These programs \citep{wu2008simple,monierEaLargeViruses08,vonMeringEaQuantitative08,stark2010mltreemap,matsen2010pplacer,berger2011performance} have various speeds and features.
This work has also spurred development of specialized alignment tools for this mapping process.
\citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
\citet{Mirarab2012-pk} use data set partitioning to improve alignments on subsets of taxa specifically for this application.
\citet{Brown2013-ia} use locality-sensitive hashing to obtain placement more than two orders of magnitude faster than the \textit{pplacer} program of \citet{matsen2010pplacer}.

@@

While developing community assembly models forms an important project for microbial ecology generally and human-associated microbial ecology in particular, phylogeny-aware methods could be further developed.
One way to model microbial community assembly is to apply Hubbell's neutral theory \citep{fierer2012animalcules,costello2012application}.
\citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
Another challenge is to bring together macroevolutionary modeling with species abundance modeling, where some initial work has been done by \citet{lambert2013predicting} in another setting.

@@

For both alpha and beta diversity measurement, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
One type of normalization handles differential depths of sequencing across samples.
The presently used approach is \textit{rarefaction}, which means uniform sub-sampling to the number of reads in the lowest abundance sample \citep{caporaso2010qiime,schloss2009introducing}.
In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another.
This is not the case.
An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
This is a good start, but more work should be done exploring results under deviation from that model.

>>>>

Despite the high throughput and low cost of modern sequencing, inherent challenges remain for applications of marker gene sequencing to take a census of microbes.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with stringent protocols, meaning that a collection of sequences need not be representative of the communities from which they were derived \citep{morgan2010metagenomic}.
Current high throughput sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
``Primer bias,'' or differing amplification levels of various sequences based on their affinity for the primers \citep{suzuki1996bias,polz1998bias}, is a challenge and has led to the standardization of primers \citep{methe2012framework}.
Worse, multiplex PCR is known to create chimeric (i.e.\ spurious recombinant) sequences via partial PCR products \citep{hugenholtz2003chimeric,ashelford2005least,haas2011chimeric,schloss2011reducing}.
Correspondingly, chimera checking software has been developed \citep[including][]{ashelford2006new,edgar2011uchime}.
Also, 16S can be present in up to 15 copies and there can be diversity within the copies \citep{klappenbach2001rrndb}.
This can distort inferences concerning actual microbe abundances based on read copy number.
Recent work by \citet{kembel2012incorporating} implements the independent contrasts method \citep{felsenstein1985phylogenies} to correct for copy number, which has been helpful despite a moderate evolutionary signal in copy number variation \citep{klappenbach2000rrna}.
Some groups have reported advantages to using alternate single-copy genes as markers for characterization of microbial communities \citep[e.g.][]{mcnabb2004hsp65,case2007rpob}, however 16S remains the dominant locus used by a large margin.
A final cause of noise is next-generation sequencing error: this is certainly a problem for both surveys and metagenomes, but is becoming less of a problem as technology improves.
I will not address it specifically except in the inference of operational taxonomic units as described below.

@@

Metagenomic data sets are often used to infer information about metabolism rather than phylogenetic nature \citep{abubucker2012metabolic,greenblum2012metagenomic}.
Discussing these methods is beyond the scope of this paper, as is the sequencing of mRNA in bulk which is called ``metatranscriptomics".

\subsection{Whole genomes}
In addition to conventional genome sequencing for microbial genomes, whole-genome sequencing from culture is currently being used for microbial outbreak tracking \citep{koser2012rapid,snitkin2012tracking}.
The Food and Drug Administration maintains GenomeTrakr, an openly accessible database of whole genomes sampled from the environment and grown in culture\footnote{\url{http://www.fda.gov/Food/FoodScienceResearch/WholeGenomeSequencingProgramWGS/}}.
This data may become more common for unculturable organisms as single-cell sequencing methods improve \citep[reviewed in][]{kalisky2011single}.
The assembly of complete genomes from metagenomes, once limited to samples with a very small number of organisms \citep{baker2010enigmatic}, is now becoming feasible for more diverse populations with improved sequencing technology and computational approaches \citep{emerson2012metagenomic,howe2012assembling,iverson2012untangling,pell2012scaling,podell2013assembly}.

@@

An obvious application of phylogenetics is to perform taxonomic classification, as the taxonomy is at least in part defined by phylogeny.
However, comparisons of taxonomic classification programs \citep{liu2008accurate,bazinet2012comparative} have indicated that current implementations of phylogenetic methods do not perform as well as simple classifiers based on counts of DNA substrings of a given length, which are often called $k$-mer classifiers \citep{wang2007naive,rosen2008metagenome}.
For those studies and others, the conventional metrics of classification performance such as precision and recall are applied to data simulated from known and taxonomically classified genomes.
% \citep{werner2011impact} Impact of training sets on classification of high-throughput bacterial 16s rRNA gene surveys
Some authors report that a combination of composition-based and homology-based classifiers work best \citep{brady2009phymm,parks2011classifying}.
The MEGAN program \citep{huson2007megan,huson2011integrative} BLASTs an unknown sequence onto a database of sequences with taxonomic labels and assigns the sequence the lowest (i.e.\ narrowest) taxonomic group shared by all of the high-quality hits; as such it is somewhat phylogenetic in that it uses the structure of the taxonomic tree.
\citet{munch2008fast,munch2008statistical} infer taxonomic assignment by automatically retrieving sequences equipped with taxonomic information and building a tree on them along with an unknown sequence.
\citet{srinivasan2012bacterial} find that phylogenetic methods to do taxonomic classification can outperform composition-based techniques at least for certain taxonomic groups.
\citet{segata2012metagenomic} propose a clever approach to inferring organisms present in a metagenomic sample by compiling a database of clade-specific genes, then classifying a given read as being from the only clade that has the corresponding gene.
They show that this has good sensitivity and specificity, however, this method can only be used to identify the presence of organisms whose genome has been sequenced.
Other metagenomic classification techniques have been reviewed by \citet{mande2012classification}; interesting recent progress has been made by \citet{Lanzen2012-ul}, \citet{Koslicki2013-mc}, and \citet{Droge2014-lj}.
\citet{treangen2013metamos} report shorter run times and much higher accuracy (again, for taxonomic classification using reads simulated from full genomes) for such metagenomic classification when reads are assembled before they are classified.

@@

There are also other methods with the less ambitious goal of mapping sequences of unknown origin into a so-called fixed reference tree, sometimes with uncertainty estimates.
These programs \citep{monierEaLargeViruses08,vonMeringEaQuantitative08,wu2008simple,matsen2010pplacer,stark2010mltreemap,berger2011performance} have various speeds and features.
This work has also spurred development of specialized alignment tools for this mapping process.
\citet{berger2011aligning} focus on the problem of inferring the optimal alignment and insertion of sequences into a tree.
\citet{Mirarab2012-pk} use data set partitioning to improve alignments on subsets of taxa specifically for this application.
\citet{Brown2013-ia} use locality-sensitive hashing to obtain placement more than two orders of magnitude faster than the \textit{pplacer} program of \citet{matsen2010pplacer}.

@@

While developing community assembly models forms an important project for microbial ecology generally and human-associated microbial ecology in particular, phylogeny-aware methods could be further developed.
One way to model microbial community assembly is to apply Hubbell's neutral theory \citep{costello2012application,fierer2012animalcules}.
\citet{o2012phylogenetic} model community assembly with an explicitly phylogenetic perspective, and include some comparison of models to data.
Continued work in this direction seems warranted, given the way in which phylogenetic tree shape statistics have had a significant impact on macroevolutionary modeling \citep{mooers1997inferring,aldous2011five}.
In this case various (alpha and beta) diversity statistics would play the role of tree shape statistics by reducing a distribution on the tips of a tree down to a real number.
Another challenge is to bring together macroevolutionary modeling with species abundance modeling, where some initial work has been done by \citet{lambert2013predicting} in another setting.

@@

For both alpha and beta diversity measurement, read count normalization has not received nearly the attention that it has in other applications of high throughput sequencing \citep[such as RNA-Seq, e.g.][]{anders2010differential,robinson2010edger}.
One type of normalization handles differential depths of sequencing across samples.
The presently used approach is \textit{rarefaction}, which means uniform sub-sampling to the number of reads in the lowest abundance sample \citep{schloss2009introducing,caporaso2010qiime}.
In addition to throwing away data, this normalization implicitly assumes a model whereby reads are sequenced independently of one another.
This is not the case.
An alternative is provided by \citet{o2012phylogenetic}, who provide a ``UniFrac score normalization curve'' based on a sampling model of community assembly.
This is a good start, but more work should be done exploring results under deviation from that model.
